---
title: "Large-scale metabolomics provides insight into natural variation in the chemical composition of oat seed"
author: "Malachy Campbell"
date: "3/26/2020"
output:
  rmdformats::html_clean:
    fig_width: 6
    fig_height: 6
    highlight: kate
    thumbnails: true
    lightbox: true
    gallery: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = "~/Documents/Dropbox/Work/Manuscripts/2020_OatLF/")
```

# Summary
This file contains all the code used to generate the results for the section titled "Large-scale metabolomics provides insight into natural variation in the chemical composition of oat seed". Below is the code to perform $k$-means clustering of accessions based on marker data, ANOVA of metabolites between subpopulations, and the code for enrichment analysis of these differentially expressed metabolites.

## Population structure in marker and metabolites

```{r, echo = T, eval = F}
library(ggplot2)
library(wesanderson)
library(cowplot)

rm(list = ls())

Input <- readRDS("GWAS/allMrks/InputData.rds")

mets <- Input$BLUPs
mrks <- Input$mrks

popStruct_Kmeans <- function(mrkData = NULL, Ks = NULL, scale = NULL){
  if(nrow(mrkData) > ncol(mrkData)){
    cat("Number of individuals (rows) is more than number of markers (columns). Maybe transpose marker matrix? \n \n")
    }
  
  if(!is.null(scale)){
    if(scale == T){
      mrkData <- scale(mrkData, center = T, scale = T)
      
      cat("Scaling and centering marker data \n \n")
    }else{
      cat("Markers won't be scaled or centered \n \n") 
      }
    }
  
  AICcalc <- function(kmodel = NULL){
    m = ncol(kmodel$centers)
    n = length(kmodel$cluster)
    k = nrow(kmodel$centers)
    D = kmodel$tot.withinss
    return(data.frame(AIC = D + 2*m*k,
                      BIC = D + log(n)*m*k))
  }
  
  kRes <- NULL
  Klist <- list()
  for(i in 1:length(Ks)){
    tmpK <- kmeans(x = mrkData, centers = Ks[i], iter.max = 10000, nstart = 100)
    kRes <- rbind(kRes, data.frame(NoK = Ks[i], 
                                   AIC = AICcalc(kmodel = tmpK)[1],
                                   BIC = AICcalc(kmodel = tmpK)[2]))
    Klist[[i]] <- tmpK
    
    cat("finished Kmeans for K = ", Ks[i], "\n \n")
  }
  
  return(list(AICs = kRes,
              Klist = Klist))
} 


noClusters <- popStruct_Kmeans(mrkData = mrks, Ks = 1:10, scale = T)
noClusters$AICs
#    NoK      AIC      BIC
# 1    1 20848464 21085127
# 2    2 19851160 20324486
# 3    3 19048272 19758260
# 4    4 18563280 19509932
# 5    5 18259162 19442477
# 6    6 17986921 19406899
# 7    7 17769584 19426224
# 8    8 17597932 19491236
# 9    9 17488407 19618373
# 10  10 17421105 19787735

mPCA <- prcomp(mets, scale. = T, center = T)
gPCA <- prcomp(mrks, scale. = T, center = T)
gPVE <- round(100*gPCA$sdev^2/sum(gPCA$sdev^2),1)
gPCs <- data.frame(Cluster = noClusters$Klist[[6]]$cluster, gPCA$x[,1:10])

mPVE <- round(100*mPCA$sdev^2/sum(mPCA$sdev^2),1)
mPCs <- data.frame(Cluster = noClusters$Klist[[6]]$cluster, mPCA$x[,1:10])

Input$subpops <- data.frame(GID = row.names(mets), cluster = mPCs$Cluster)
saveRDS(Input, "GWAS/allMrks/InputData.rds")

genP12 <- ggplot(gPCs, aes(x = PC1, y = PC2)) +
          geom_point(aes(fill = factor(Cluster)), 
                     colour = "black", pch = 21, size = 1.5) + 
          scale_fill_manual(name = "Cluster",
                            values = c(wes_palette("Darjeeling2"), 
                                       wes_palette("Darjeeling1")[2])) +
          labs(x = paste0("PC1 (PVE: ", round(gPVE[1],1), "%)"), 
               y = paste0("PC2 (PVE: ", round(gPVE[2],1), "%)")) + 
          theme_bw() + 
          theme(plot.background = element_blank(),
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(), 
                aspect.ratio = 1, 
                legend.title = element_text(colour = "black", size = 8), 
                axis.text.x = element_text(colour = "black", size = 6), 
                axis.text.y = element_text(colour = "black", size = 6),
                axis.title = element_text(size = 8)) +
          theme(plot.margin=unit(c(1.5,2,1,1.5),"pt")) #t,r,b,l

genP34 <- ggplot(gPCs, aes(x = PC3, y = PC4)) +
          geom_point(aes(fill = factor(Cluster)), 
                     colour = "black", pch = 21, size = 1.5) +  
          scale_fill_manual(name = "Cluster",
                            values = c(wes_palette("Darjeeling2"), 
                                       wes_palette("Darjeeling1")[2])) +
          labs(x = paste0("PC3 (PVE: ", round(gPVE[3],1), "%)"), 
               y = paste0("PC4 (PVE: ", round(gPVE[4],1), "%)")) + 
          theme_bw() + 
          theme(plot.background = element_blank(),
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(), 
                aspect.ratio = 1, 
                legend.title = element_text(colour = "black", size = 8), 
                axis.text.x = element_text(colour = "black", size = 6), 
                axis.text.y = element_text(colour = "black", size = 6),
                axis.title = element_text(size = 8)) +
          theme(plot.margin=unit(c(1.5,2,1,1.5),"pt")) #t,r,b,l

metP12 <- ggplot(mPCs, aes(x = PC1, y = PC2)) +
          geom_point(aes(fill = factor(Cluster)), 
                     colour = "black", pch = 21, size = 1.5) +  
          scale_fill_manual(name = "Cluster",
                            values = c(wes_palette("Darjeeling2"), 
                                       wes_palette("Darjeeling1")[2])) +
          labs(x = paste0("PC1 (PVE: ", round(mPVE[1],1), "%)"), 
               y = paste0("PC2 (PVE: ", round(mPVE[2],1), "%)")) + 
          theme_bw() + 
          theme(plot.background = element_blank(),
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(), 
                aspect.ratio = 1, 
                legend.title = element_text(colour = "black", size = 8), 
                axis.text.x = element_text(colour = "black", size = 6), 
                axis.text.y = element_text(colour = "black", size = 6),
                axis.title = element_text(size = 8)) +
          theme(plot.margin=unit(c(1.5,2,1,1.5),"pt")) #t,r,b,l

metP34 <- ggplot(mPCs, aes(x = PC3, y = PC4)) +
          geom_point(aes(fill = factor(Cluster)), 
                     colour = "black", pch = 21, size = 1.5) + 
          scale_fill_manual(name = "Cluster",
                            values = c(wes_palette("Darjeeling2"), 
                                       wes_palette("Darjeeling1")[2])) +
          labs(x = paste0("PC3 (PVE: ", round(mPVE[3],1), "%)"), 
               y = paste0("PC4 (PVE: ", round(mPVE[4],1), "%)")) + 
          theme_bw() + 
          theme(plot.background = element_blank(),
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(), 
                aspect.ratio = 1, 
                legend.title = element_text(colour = "black", size = 8), 
                axis.text.x = element_text(colour = "black", size = 6), 
                axis.text.y = element_text(colour = "black", size = 6),
                axis.title = element_text(size = 8)) +
          theme(plot.margin=unit(c(1.5,2,1,1.5),"pt")) #t,r,b,l

Pgrid <- plot_grid(genP12 + theme(legend.position = "none"), 
                   genP34 + theme(legend.position = "none"), 
                   metP12 + theme(legend.position = "none"), 
                   metP34 + theme(legend.position = "none"),
                   ncol = 2, nrow = 2,
                   labels = c("A", "B", "C", "D"), label_size = 8, 
                   hjust = -0.5, vjust = 1.2)

legs <- get_legend(genP12) #get the legend

pdf("Figs/PCA_allMrks.pdf", h = 5.4, w = 5.4)
plot_grid(Pgrid, legs, rel_widths = c(2, 0.2))
dev.off()
```




## Do metabolites differ between subpopulations?
```{r, echo = T, eval = F}
library(car)

rm(list = ls())

Input <- readRDS("flashr/GWAS/allMrks/InputData.rds") #note this analysis uses only 335 accessions with marker data
mets <- Input$BLUPs
subpops <- Input$subpops

sum(row.names(mets) == subpops$GID)

AOVres <- NULL
for (i in 1:ncol(mets)){
  tmp.df <- data.frame(GID = subpops$GID, subpop = as.factor(subpops$cluster), Y = mets[,i])
  tmp.aov <- Anova(lm(Y ~ subpop, data = tmp.df), type="III")
  AOVres <- rbind(AOVres, 
                  data.frame(cmpd = colnames(mets)[i], term = row.names(tmp.aov), tmp.aov))
  }
AOVres <- AOVres[AOVres$term %in% "subpop" ,]
AOVres$padj <- p.adjust(AOVres$Pr..F., method = "BH")

dim(AOVres[AOVres$padj < 0.01 ,])[1]/nrow(AOVres) #0.407

write.csv(AOVres, "SuppFiles/subpopAOV.csv", row.names = F)
```

### Enrichment analysis
Do specific pathways/functional classes vary between subpopulations?
```{r, echo = T, eval = F}
library(xtable)

rm(list = ls())

###Function for enrichment test using a hypergeometric test
hypergeoTest <- function(Annots = NULL, AllCmpdVect = NULL, SigCmpdVect = NULL, AnnotClass = NULL, minNoCmpds = NULL){
  #Annots: dataframe of annotations with a column called cmpd and the rest of the columns indicating an annotation group/class; same format as trevor's files
  #AllCmpdVect: a vector of compound names. This is the population of compounds that were used for analysis
  #SigCmpdVect: a vector of compound names that were significantly different between groups/subpops
  #AnnotClass: a string that should match a column name in Annots; determines the level of the heirarchy to perform the test(s)
  #minNoCmpds: min number of compounds in a given class. Test will not be run if no in class is < minNoCmpds
  
  tmpAnnot <- Annots[c("cmpd", AnnotClass)]
  tmpAnnot <- na.omit(tmpAnnot) #remove compounds with no annotation
  tmpAnnot <- tmpAnnot[tmpAnnot$cmpd %in% AllCmpdVect ,] #only use metabolites that were considered for analysis
  
  #Only metabolites with annotations
  AllCmpdVect <- AllCmpdVect[AllCmpdVect %in% tmpAnnot$cmpd]
  SigCmpdVect <- SigCmpdVect[SigCmpdVect %in% tmpAnnot$cmpd]
  
  colnames(tmpAnnot) <- c("cmpd", "class")
  
  #x: Number of annotated, differentially expressed metabolites in a given class (number of white balls drawn)
  #m: Number of annotated metabolites belonging to a given class (total number of white balls)
  #n: Number of annotated metabolites NOT belonging to a given class (total number of non-white balls)
  #k: Number of annotated metabolites that are differentially expressed (number of balls drawn)
  enrichRes <- NULL
  
  for(i in 1:length(unique(tmpAnnot$class))){
    tmpClass <- unique(tmpAnnot$class)[i]
    tmptmpAnnot <- tmpAnnot[tmpAnnot$class %in% tmpClass ,] #extract all compounds in given annotation class
    
    K = length(SigCmpdVect)
    X = length(SigCmpdVect[SigCmpdVect %in% tmptmpAnnot$cmpd])
    N = length(AllCmpdVect) - X
    M = nrow(tmptmpAnnot)
    
    if(is.null(minNoCmpds)){
      enrichRes <- rbind(enrichRes, 
                       data.frame(heirLevel = AnnotClass,
                                  annGroup = tmpClass,
                                  NoDE_k = K,
                                  NoDEinClass_x = X,
                                  NoInClass_m = M,
                                  NoWithAnn = length(AllCmpdVect),
                                  pVal = phyper(q = X, m = M, n = N, k = K, lower.tail = F)))
    }else{
      if(M < minNoCmpds){
        enrichRes <- rbind(enrichRes, 
                       data.frame(heirLevel = AnnotClass,
                                  annGroup = tmpClass,
                                  NoDE_k = K,
                                  NoDEinClass_x = X,
                                  NoInClass_m = M,
                                  NoWithAnn = length(AllCmpdVect),
                                  pVal = NA))
      }else{
        enrichRes <- rbind(enrichRes, 
                       data.frame(heirLevel = AnnotClass,
                                  annGroup = tmpClass,
                                  NoDE_k = K,
                                  NoDEinClass_x = X,
                                  NoInClass_m = M,
                                  NoWithAnn = length(AllCmpdVect),
                                  pVal = phyper(q = X, m = M, n = N, k = K, lower.tail = F)))
      }
    }
  }
  enrichRes$AdjP <- p.adjust(enrichRes$pVal, method = "BH")
  return(enrichRes)
}
#####

#Load annotations
Annot.GC <- read.csv("flashr/metaFiles/GCMS_2018_GC.03_annotation.csv")[c(1,3,4,9:14)]
Annot.LC <- read.csv("flashr/metaFiles/LCMS_2018_LC.03_annotation.csv")[c(1,3,4,16:21)]
Annot <- rbind(Annot.GC, Annot.LC)
rm(Annot.LC, Annot.GC)

#Load AOV res
AOVres <- read.csv("SuppFiles/subpopAOV.csv")
AOVresSig <- AOVres[AOVres$padj < 0.01 ,]


enrichKingdom <- hypergeoTest(Annots = Annot, AllCmpdVect = AOVres$cmpd, SigCmpdVect = AOVresSig$cmpd, AnnotClass = "kingdom", minNoCmpds = 3)
enrichSuperClass <- hypergeoTest(Annots = Annot, AllCmpdVect = AOVres$cmpd, SigCmpdVect = AOVresSig$cmpd, AnnotClass = "superclass", minNoCmpds = 3)
enrichClass <- hypergeoTest(Annots = Annot, AllCmpdVect = AOVres$cmpd, SigCmpdVect = AOVresSig$cmpd, AnnotClass = "class", minNoCmpds = 3)
enrichSubClass <- hypergeoTest(Annots = Annot, AllCmpdVect = AOVres$cmpd, SigCmpdVect = AOVresSig$cmpd, AnnotClass = "subclass", minNoCmpds = 3)
enrichParent <- hypergeoTest(Annots = Annot, AllCmpdVect = AOVres$cmpd, SigCmpdVect = AOVresSig$cmpd, AnnotClass = "parent", minNoCmpds = 3)

AllRes <- rbind(enrichKingdom, enrichSuperClass, enrichClass, enrichSubClass, enrichParent)
xtable(na.omit(AllRes)[na.omit(AllRes)$AdjP <= 0.07 ,][c("heirLevel", "annGroup", "NoDEinClass_x", "NoInClass_m", "pVal", "AdjP")])

write.csv(AllRes,
          "flashr/manuscriptFigs/DE_EnrichmentTest.csv", row.names = F)
```